Automatically proving equivalence by type-safe reflection

### 2016-01-18 paper-JFP-DTP/content/4-detailsNormalisation.tex
Section 4 finished

\section {Normalisation functions and re-usability of the provers}

@@

\subsection {Normalization of terms in a Monoid}

@@

%\begin{figure}[H]
%\figrule
%\begin{center}
%\begin{lstlisting}
%elimZero : (p:Monoid c)->(gam:Vect n c)->{c1:c} 
%            -> (ExprMo p gam c1) 
%            -> (c2 ** (ExprMo p gam c2, c1 $\simeq$ c2))
%elimZero c p gam (PlusMo (ConstMo _ _ const1) (VarMo _ v)) 
%        with (eq_on_set p Zero const1)
%   elimZero c p gam (PlusMo (ConstMo _ _ Zero) (VarMo _ v)) 
%        | (Just refl) = (_ ** (VarMo _ v, ?MelimZero1))
%   elimZero c p gam (PlusMo (ConstMo _ _ const1) (VarMo _ v)) 
%        | _ = (_ ** (PlusMo (ConstMo _ _ const1) (VarMo _ v), refl)) 
%elimZero c p gam 
% (PlusMo (VarMo _ v) (ConstMo _ _ const2)) 
%      with (eq_as_elem_of_set p Zero const2) 
%   elimZero c p gam (PlusMo (VarMo _ v) 
%            (ConstMo _ _ Zero)) | (Just refl) 
%          = (_ ** (VarMo _ _ v, ?MelimZero2))
%   elimZero c p gam (PlusMo (VarMo _ v) 
%            (ConstMo _ _ const2)) | _ 
%          = (_ ** (PlusMo (VarMo _ v) 
%                 (ConstMo _ _ const2), refl))
%[...]
%\end{lstlisting}
%\end{center}
%\caption{Eliminating zeros in a Monoid, first pattern}
%\figrule
%\end{figure}

@@

\subsection {Reusing the monoid prover for building the group prover}

@@

The idea : negations $-x$ are encoded as variables.

@@

The idea : monomials (containing products), like $3*(x*y)$ are encoded as variables when passed to the commutative group prover.

>>>>

\section{Normalisation functions and Reusability of Provers}

\label{sect:reuse}

@@

\subsection {Reusing the semigroup prover for building the monoid prover}

@@

@@

@@

The last remaining step is to call the monoid prover. Unfortunately, we can't directly write a function of conversion from group to monoid, because in a group we have the possibility to express negations and subtractions, that we do not have in a monoid. The idea is that we will encode negations as variables, and we will let the monoid prover deal with them as ordinary variables. In order to achieve this goal, we need the following datatype :

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data Variable : {c:Type} -> {n:Nat} -> (Vect n c) -> c -> Type where
    RealVariable : ($\Gamma$:Vect n c) -> (i:Fin n) -> Variable $\Gamma$ (index i $\Gamma$)
    EncodingNegOfVar : ($\Gamma$:Vect n c) -> (i:Fin n) -> Variable $\Gamma$ (Neg (index i $\Gamma$))
\end{lstlisting}
\end{center}
\caption{Real variables and encodings}
\figrule
\end{figure}

We only need to encode negations of variables, as negations of constants can be simplified into a constant. Also, there can't be a negation of something more complicated than an atom (a variable or a constant), because of the treatments we've done beforehand, especially because we've propagated all the negations symbols inward the parenthesis.

The constructor for variable now takes a Variable, instead of taking directly a $(Fin\ n)$.
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
    VarG : (p:Group c) -> {$\Gamma$:Vect n c} -> {val:c} 
          -> (Variable $\Gamma$ val) -> ExprG p $\Gamma$ val
\end{lstlisting}
\end{center}
\caption{new Var constructor}
\figrule
\end{figure}

Thanks to this encoding, we can now transform an $ExprG$ to an $ExprMo$. A constant $(ConstG\ p\ \Gamma\ c1)$ will be transformed into the corresponding constant $(ConstMo\ p\ \Gamma\ c1)$, a $PlusG$ into the corresponding $PlusMo$, a variable $(VarG\ p\ var)$ into the corresponding variable $(VarMo\ p\ var)$, the negation of a constant into the resulting constant, and finally the negation of a variable $i$ into a $(Var\ p\ \Gamma\ (EncodingNegOfVar\ \Gamma\ i))$.
Once this encoding is computed, the antepenultimate step is to call the monoid prover and to let it finish the normalisation. Finally, we need to convert back the result into an $ExprG$ by interpreting the $EncodingNegOfVar$ into a $Neg$ of the corresponding variable.

\subsection {Reusing the group prover for building the commutative group prover}

The normalisation of the commutative group prover starts by calling the group prover. Here, the two algebraic structures have the same power of expressivity, and thus there's no specific encoding to design. Then, we simply need to decide an ordering for variables and constants. We arbitrary decide that $Var\ FZ\ <\ Var\ (FS\ FZ)\ <\ Var\ (FS\ (FS\ FZ))\ <\ ...\ <\ Constant$. The normalisation function will therefore simply sort the expression with this ordering and will do basic simplifications.

@@

The most important specific task of the ring reduction is to develop entirely the expression with the use of distributivity. After that, it needs to call the commutative group prover, but again, we're losing some expressivity as it won't be possible to express a product at the commutative group level. We follow the same idea that we've implemented for the conversion going from group to monoid, and we will use some specific encoding. More precisely, all monomials (containing products), like $3*(x*y)$ are encoded as variables when passed to the commutative group prover. This is possible because we've fully developed the expression, which had for effect to ensure that a product can't contain a sum.
We will add a coonstructor to the type $Variable$ in order to encode these monomials.

### 2016-01-18 paper-JFP-DTP/content/5-results.tex
Section 4 finished

In a context we're we've got the following induction hypothesis \[ihn:(plus\ (plus\ c\ bit)\ bit1\ =\ plus\ vLsb\ (plus\ vCarry0\ (plus\ vCarry0\ 0)))\]
The goal can't be solved directly by calling commutative monoid prover, as it uses different variables in its left and right hand side. However, the induction hypothesis $ihn$ --obtained with the recursive call-- creates a link between these variables. Therefore, we can use our reduction procedure for commutative monoid four times : two for reducing the LHS and RHS of $ihn$, and two for reducing the LHS and RHS of the goal. After these four normalisations, it becomes possible to simply rewrite the new induction impothesis $ihn'$ on the current goal.

@@

This example was a bit unconventional in the sense that 4 calls to the normalisation function were needed. Usually, only two calls to the normalisation are enough, one for the left hand side and one for the right hand side, and they are done in the decision procedure $commutativeMonoidDecideEq$. These two extra calls were needed because of the known equality that we had from the induction hypothesis. This example\footnote{Which can be found in $RingIdris/Examples/binary.idr$}, which was therefore at the boundary to deduction modulo [REF], shows that our implementation can deal with concrete goals obtained during the development of real programs.

>>>>

In a context we're we've got the following induction hypothesis \[ihn:(plus\ (plus\ c\ bit)\ bit1\ =\ plus\ vLsb\ (plus\ vCarry0\ (plus\ vCarry0\ 0)))\]
The goal can't be solved directly by calling the commutative monoid prover, as it uses different variables in its left and right hand side. However, the induction hypothesis $ihn$ --obtained with the recursive call-- creates a link between these variables. Therefore, the solution is simply to use our reduction procedure for commutative monoid four times : two for reducing the LHS and RHS of $ihn$ --that gives a new proof of equality $ihn'$--, and two for reducing the LHS and RHS of the goal. After these four normalisations, it becomes possible to simply rewrite the new induction impothesis $ihn'$ on the current goal.

@@

This example was a bit unconventional in the sense that 4 calls to the normalisation function were needed. Usually, only two calls to the normalisation are enough, one for the left hand side and one for the right hand side, and they are done automatically in the decision procedure $commutativeMonoidDecideEq$. These two extra calls were needed because of the known equality that we had from the induction hypothesis. This example\footnote{Which can be found in $RingIdris/Examples/binary.idr$}, which was therefore at the boundary to deduction modulo~\cite{Dowek03}, shows that our implementation can deal with concrete goals obtained during the development of real programs.
